---
title: "STL for Great Lakes Temperature Data"
author: "Song Qian"
date: "8/1/2019"
output: pdf_document
editor_options:
  chunk_output_type: console
---

# Introduction
This document includes R code for analyzing GL temperature data using STL.  Because the current implementation of STL in R does not allow missing values, we used the median polishing algorithm to impute missings before using STL. The function `median.polish.ts()` replaces missing monthly temperature with the estimated annual and monthly median. It also produces two figures to show the seasonal and de-seasoanlized trends.  We also modified the STL function to produce desired figures using the function `stl.rfs2()`.  This function is a wrapper of the R function `stl()`. It produces the figure and returns data used for the figure. The R code chunk is not included in the knitted PDF file.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy=TRUE)

packages<-function(x, repos="http://cran.r-project.org", ...){
  x<-as.character(match.call()[[2]])
  if (!require(x,character.only=TRUE)){
    install.packages(pkgs=x, repos=repos, ...)
    require(x,character.only=TRUE)
  }
}

median.polish.ts <- function(data.ts, ylab="", plt=T){
  # median polishing for missing value imputation
  medpolish(matrix(data.ts, ncol=12, byrow=T), eps=0.001, na.rm=T)->temp.2w
  print(names(temp.2w))
  year.temp <- rep(seq(start(data.ts)[1], end(data.ts)[1]), each=12)
  month.temp <- rep(1:12, length(seq(start(data.ts)[1], end(data.ts)[1])))
  # plotting median polishing results
  if (plt){
    par(mfrow=c(2,1))
    plot(seq(start(data.ts)[1], end(data.ts)[1]),
         temp.2w$overall+temp.2w$row, type="l",
         xlab="Year", ylab=ylab, main="De-seasonalized Trend")
    plot(seq(1,12), temp.2w$overall+temp.2w$col, type="l",
         xlab="Month", ylab=ylab, main="Seasonal Changes")
  }
  data.ts[is.na(data.ts)]<-temp.2w$overall +
    temp.2w$row[year.temp[is.na(data.ts)]-start(data.ts)[1]+1]+
      temp.2w$col[month.temp[is.na(data.ts)]]
  data.ts[is.na(data.ts)]<-temp.2w$overall
  invisible(data.ts)
}

stl.rfs2 <-
    function(data = carbon.dioxide, ss.w = 25, ss.d = 1, fc.w = 120, 
             fc.d = 1, ylab = "Carbon Dioxide (ppm)", aspect = "xy", ...)
{
    strip.background <- trellis.par.get("strip.background")
    strip.background$col <- 0.
    trellis.par.set("strip.background", strip.background)
    strip.shingle <- trellis.par.get("strip.shingle")
    strip.shingle$col <- 0.
    trellis.par.set(theme = canonical.theme("postscript", col=FALSE))
    trellis.par.set("strip.shingle", strip.shingle)
    the.fit <- stl(data, s.window = ss.w, s.degree = ss.d, t.window = fc.w,
                   t.degree = fc.d, ...)
    sfit <- the.fit$time.series[,1]
    tfit <- the.fit$time.series[,2]
    fit.time <- time(data)
    car.subseries <- factor(cycle(data), label = month.abb)
    obj1.df <- data.frame(y=sfit, x=fit.time, cat=car.subseries)

    obj1 <- xyplot(sfit ~ fit.time | car.subseries, layout = c(12, 1), panel = function(x, y)
    {
        panel.xyplot(x, y, type = "l")
        panel.abline(h = mean(y))
    }
  , aspect = aspect, xlab = "Year", ylab = ylab)

    obj2.df <- data.frame(y=tfit, x=fit.time)
    obj2 <- xyplot(tfit ~ fit.time, panel = function(x, y)
        panel.xyplot(x, y, type = "l"), xlab = "", aspect = "xy", ylab = "")

    n <- length(data)
    the.fit.trend <- the.fit$time.series[,2] - mean(the.fit$time.series[,2])
    fit.components <- c(the.fit.trend, the.fit$time.series[,1], the.fit$time.series[,3])
    fit.time <- rep(time(data), 3)
    fit.names <- ordered(rep(c("Trend", "Seasonality", "Residuals"), c(n, n, n)),
                         c("Trend", "Seasonality", "Residuals"))

    obj3.df <- data.frame(x=fit.time, y=fit.components, series=fit.names)
    obj3 <- xyplot(fit.components ~ fit.time | fit.names, panel = function(x, y)
    {
        panel.grid(h = 5)
        panel.xyplot(x, y, type = "l")
    }, aspect=0.75, layout = c(3, 1), ylim = c(-1, 1) * max(abs(fit.components)),
    xlab = "", ylab = ylab)
    print(obj1, position = c(0, 0, 1, 0.6), more = T)
    print(obj3, position = c(0, 0.4, 1, 1), more = F)
    return(list(obj1.df, obj2.df, obj3.df, data.frame(trend=the.fit$time.series[,2],
                         season=the.fit$time.series[,1],
                         residual=the.fit$time.series[,3])))
}
```

# Setting Up File System
The following R code chunk defines the default file systems: all input data are in the subfolder `Data`, and resulting figures are saved in subfolder `Figs`.  

```{r}
base <- getwd()  ##"~/Google Drive/UT/Research/CILER/STL"
dataDIR <- paste(base, "Data", sep="/")
figDIR <- paste(base, "Figs", sep="/")
dataOUTDIR <- paste(dataDIR, "output", sep="/")

packages(tidyverse)
packages(readxl)
packages(lattice)
```

# Reading Data

The following R code chunk imports data, processes missing values, and creates time series objects.

```{r}
stl_data <- read_excel(paste(dataDIR, "glsea_trends.xlsx", sep="/"),
                       col_names=F)
stl_data_therm <- read_excel(paste(dataDIR, "thermistor_trends.xlsx", sep="/"),
                             col_names=F)
Superior <- matrix(as.numeric(unlist(stl_data[4:27,-1])), ncol=12)
Michigan <- matrix(as.numeric(unlist(stl_data[30:53,-1])), ncol=12)
Huron <- matrix(as.numeric(unlist(stl_data[56:79,-1])), ncol=12)
StClair <- matrix(as.numeric(unlist(stl_data[82:105,-1])), ncol=12)
Erie <- matrix(as.numeric(unlist(stl_data[108:131,-1])), ncol=12)
Ontario <- matrix(as.numeric(unlist(stl_data[134:157,-1])), ncol=12)

therm_surf <- matrix(as.numeric(unlist(stl_data_therm[4:27, -1])), ncol=12)
therm_25m  <- matrix(as.numeric(unlist(stl_data_therm[30:58, -1])), ncol=12)
therm_50m  <- matrix(as.numeric(unlist(stl_data_therm[60:89, -1])), ncol=12)
therm_75m  <- matrix(as.numeric(unlist(stl_data_therm[92:120, -1])), ncol=12)
therm_100m <- matrix(as.numeric(unlist(stl_data_therm[123:151, -1])), ncol=12)
therm_125m <- matrix(as.numeric(unlist(stl_data_therm[154:182, -1])), ncol=12)

therm_surf[therm_surf< -900] <- NA
therm_25m[therm_25m < -900] <- NA
therm_50m[therm_50m < -900] <- NA
therm_75m[therm_75m < -900] <- NA
therm_100m[therm_100m < -900] <- NA
therm_125m[therm_125m < -900] <- NA

SP_monthly <- ts(as.vector(t(Superior)), start=c(1995,1), end=c(2018,12), frequency=12)
MI_monthly <- ts(as.vector(t(Michigan)), start=c(1995,1), end=c(2018,12), frequency=12)
HR_monthly <- ts(as.vector(t(Huron)), start=c(1995,1), end=c(2018,12), frequency=12)
SC_monthly <- ts(as.vector(t(StClair)), start=c(1995,1), end=c(2018,12), frequency=12)
ER_monthly <- ts(as.vector(t(Erie)), start=c(1995,1), end=c(2018,12), frequency=12)
ON_monthly <- ts(as.vector(t(Ontario)), start=c(1995,1), end=c(2018,12), frequency=12)

thermSurf_monthly <- ts(as.vector(t(therm_surf)), start=c(1990, 1), frequency=12)
therm25m_monthly <- ts(as.vector(t(therm_25m)), start=c(1990, 1), frequency=12)
therm50m_monthly <- ts(as.vector(t(therm_50m)), start=c(1990, 1), frequency=12)
therm75m_monthly <- ts(as.vector(t(therm_75m)), start=c(1990, 1), frequency=12)
therm100m_monthly <- ts(as.vector(t(therm_100m)), start=c(1990, 1), frequency=12)
therm125m_monthly <- ts(as.vector(t(therm_125m)), start=c(1990, 1), frequency=12)
```

# STL Analysis

The analysis uses the function `stl.rfs2()` to produce the STL plots for each lake and export data for reproducing the figures using other graphical systems.

```{r}
pdf(paste(figDIR, "SuperiorSTL.pdf", sep="/"), height=4, width=5)
Superior.STL <- stl.rfs2(data=SP_monthly,
                   aspect=2.25, ylab="Temperature",
                   ss.w = 11, ss.d = 1, fc.w = 91, fc.d = 1)
dev.off()

write.csv(Superior.STL[[3]], paste(dataOUTDIR, "superior_monthly.csv", sep="/"))
write.csv(Superior.STL[[1]], paste(dataOUTDIR, "superior_components.csv", sep="/"))

pdf(paste(figDIR, "MichiganSTL.pdf", sep="/"), height=4, width=5)
Michigan.STL <- stl.rfs2(data=MI_monthly,
                   aspect=2.25, ylab="Temperature",
                   ss.w = 15, ss.d = 1, fc.w = 91, fc.d = 1)
dev.off()

write.csv(Michigan.STL[[3]], paste(dataOUTDIR, "michigan_monthly.csv", sep="/"))
write.csv(Michigan.STL[[1]], paste(dataOUTDIR, "michigan_components.csv", sep="/"))

pdf(paste(figDIR, "HuronSTL.pdf", sep="/"), height=4, width=5)
Huron.STL <- stl.rfs2(data=HR_monthly,
                   aspect=2.25, ylab="Temperature",
                   ss.w = 15, ss.d = 1, fc.w = 91, fc.d = 1)
dev.off()

write.csv(Huron.STL[[3]], paste(dataOUTDIR, "huron_monthly.csv", sep="/"))
write.csv(Huron.STL[[1]], paste(dataOUTDIR, "huron_components.csv", sep="/"))

pdf(paste(figDIR, "ErieSTL.pdf", sep="/"), height=4, width=5)
Erie.STL <- stl.rfs2(data=ER_monthly,
                   aspect=2.25, ylab="Temperature",
                   ss.w = 15, ss.d = 1, fc.w = 91, fc.d = 1)
dev.off()

write.csv(Erie.STL[[3]], paste(dataOUTDIR, "erie_monthly.csv", sep="/"))
write.csv(Erie.STL[[1]], paste(dataOUTDIR, "erie_components.csv", sep="/"))

pdf(paste(figDIR, "StClairSTL.pdf", sep="/"), height=4, width=5)
StClair.STL <- stl.rfs2(data=SC_monthly,
                   aspect=2.25, ylab="Temperature",
                   ss.w = 15, ss.d = 1, fc.w = 91, fc.d = 1)
dev.off()

write.csv(StClair.STL[[3]], paste(dataOUTDIR, "stclair_monthly.csv", sep="/"))
write.csv(StClair.STL[[1]], paste(dataOUTDIR, "stclair_components.csv", sep="/"))

pdf(paste(figDIR, "OntarioSTL.pdf", sep="/"), height=4, width=5)
Ontario.STL <- stl.rfs2(data=ON_monthly,
                   aspect=2.25, ylab="Temperature",
                   ss.w = 15, ss.d = 1, fc.w = 91, fc.d = 1)
dev.off()

write.csv(Ontario.STL[[3]], paste(dataOUTDIR, "ontario_monthly.csv", sep="/"))
write.csv(Ontario.STL[[1]], paste(dataOUTDIR, "ontario_components.csv", sep="/"))

therm25_ts <- median.polish.ts(therm25m_monthly)
pdf(paste(figDIR, "Term_25mSTL.pdf", sep="/"), height=4.75, width=5)
Therm_25m.STL <- stl.rfs2(data=therm25_ts,
                   aspect=2.25, ylab="Temperature",
                   ss.w = 15, ss.d = 1, fc.w = 91, fc.d = 1)
dev.off()

write.csv(Therm_25m.STL[[3]], paste(dataOUTDIR, "therm25m_monthly.csv", sep="/"))
write.csv(Therm_25m.STL[[1]], paste(dataOUTDIR, "therm25m_components.csv", sep="/"))

thermsurf_ts <- median.polish.ts(thermSurf_monthly)
pdf(paste(figDIR, "Term_surfSTL.pdf", sep="/"), height=4.75, width=5)
Therm_surf.STL <- stl.rfs2(data=thermsurf_ts,
                   aspect=2.25, ylab="Temperature",
                   ss.w = 15, ss.d = 1, fc.w = 91, fc.d = 1)
dev.off()

write.csv(Therm_surf.STL[[3]], paste(dataOUTDIR, "thermsurf_monthly.csv", sep="/"))
write.csv(Therm_surf.STL[[1]], paste(dataOUTDIR, "thermsurf_components.csv", sep="/"))

therm75_ts <- median.polish.ts(therm75m_monthly)
pdf(paste(figDIR, "Term_75mSTL.pdf", sep="/"), height=4.75, width=5)
Therm_75m.STL <- stl.rfs2(data=therm75_ts,
                   aspect=2.25, ylab="Temperature",
                   ss.w = 15, ss.d = 1, fc.w = 91, fc.d = 1)
dev.off()

write.csv(Therm_75m.STL[[3]], paste(dataOUTDIR, "therm75m_monthly.csv", sep="/"))
write.csv(Therm_75m.STL[[1]], paste(dataOUTDIR, "therm75m_components.csv", sep="/"))

therm50_ts <- median.polish.ts(therm50m_monthly)
pdf(paste(figDIR, "Term_50mSTL.pdf", sep="/"), height=4.75, width=5)
Therm_50m.STL <- stl.rfs2(data=therm50_ts,
                   aspect=2.25, ylab="Temperature",
                   ss.w = 15, ss.d = 1, fc.w = 91, fc.d = 1)
dev.off()

write.csv(Therm_50m.STL[[3]], paste(dataOUTDIR, "therm50m_monthly.csv", sep="/"))
write.csv(Therm_50m.STL[[1]], paste(dataOUTDIR, "therm50m_components.csv", sep="/"))

therm100_ts <- median.polish.ts(therm100m_monthly)
pdf(paste(figDIR, "Term_100mSTL.pdf", sep="/"), height=4.75, width=5)
Therm_100m.STL <- stl.rfs2(data=therm100_ts,
                   aspect=2.25, ylab="Temperature",
                   ss.w = 15, ss.d = 1, fc.w = 91, fc.d = 1)
dev.off()

write.csv(Therm_100m.STL[[3]], paste(dataOUTDIR, "therm100m_monthly.csv", sep="/"))
write.csv(Therm_100m.STL[[1]], paste(dataOUTDIR, "therm100m_components.csv", sep="/"))

therm125_ts <- median.polish.ts(therm125m_monthly)
pdf(paste(figDIR, "Term_125mSTL.pdf", sep="/"), height=4.75, width=5)
Therm_125m.STL <- stl.rfs2(data=therm125_ts,
                   aspect=2.25, ylab="Temperature",
                   ss.w = 15, ss.d = 1, fc.w = 91, fc.d = 1)
dev.off()

write.csv(Therm_125m.STL[[3]], paste(dataOUTDIR, "therm125m_monthly.csv", sep="/"))
write.csv(Therm_125m.STL[[1]], paste(dataOUTDIR, "therm125m_components.csv", sep="/"))
```

## New data
This R code chunk was written to process the updated data.
```{r}
stl_data2 <- read_excel(paste(dataDIR, "updated_temperature_depth_data.xlsx",
                              sep="/"),  col_names=F)
Surface <- matrix(as.numeric(unlist(stl_data2[3:26,-c(1,2)])), ncol=12)
D30m <- matrix(as.numeric(unlist(stl_data2[30:58,-c(1,2)])), ncol=12)
D60m <- matrix(as.numeric(unlist(stl_data2[62:90,-c(1,2)])), ncol=12)
D75m <- matrix(as.numeric(unlist(stl_data2[94:122,-c(1,2)])), ncol=12)
D100 <- matrix(as.numeric(unlist(stl_data2[126:154,-c(1,2)])), ncol=12)
D110 <- matrix(as.numeric(unlist(stl_data2[158:186,-c(1,2)])), ncol=12)
D120 <- matrix(as.numeric(unlist(stl_data2[190:218,-c(1,2)])), ncol=12)
D140 <- matrix(as.numeric(unlist(stl_data2[222:250,-c(1,2)])), ncol=12)

Surface[Surface< -900] <- NA
D30m[D30m < -900] <- NA
D60m[D60m < -900] <- NA
D75m[D75m < -900] <- NA
D100[D100 < -900] <- NA
D110[D110 < -900] <- NA
D120[D120 < -900] <- NA
D140[D140 < -900] <- NA

SF_monthly <- median.polish.ts(ts(as.vector(t(Surface)), start=c(1995,1), end=c(2018,12), frequency=12))
D30m_monthly <- median.polish.ts(ts(as.vector(t(D30m)), start=c(1990,1), end=c(2018,12), frequency=12))
D60m_monthly <- median.polish.ts(ts(as.vector(t(D60m)), start=c(1990,1), end=c(2018,12), frequency=12))
D75m_monthly <- median.polish.ts(ts(as.vector(t(D75m)), start=c(1990,1), end=c(2018,12), frequency=12))
D100_monthly <- median.polish.ts(ts(as.vector(t(D100)), start=c(1990,1), end=c(2018,12), frequency=12))
D110_monthly <- median.polish.ts(ts(as.vector(t(D110)), start=c(1990,1), end=c(2018,12), frequency=12))
D120_monthly <- median.polish.ts(ts(as.vector(t(D120)), start=c(1990,1), end=c(2018,12), frequency=12))
D140_monthly <- median.polish.ts(ts(as.vector(t(D140)), start=c(1990,1), end=c(2018,12), frequency=12))

pdf(paste(figDIR, "SurfaceSTL.pdf", sep="/"), height=4, width=5)
Surface.STL <- stl.rfs2(data=SF_monthly,
                   aspect=2.25, ylab="Temperature",
                   ss.w = 11, ss.d = 1, fc.w = 91, fc.d = 1)
dev.off()

write.csv(Surface.STL[[3]], paste(dataOUTDIR, "surface_monthly.csv", sep="/"))
write.csv(Surface.STL[[1]], paste(dataOUTDIR, "surface_components.csv", sep="/"))

pdf(paste(figDIR, "D30mSTL.pdf", sep="/"), height=4, width=5)
D30m.STL <- stl.rfs2(data=D30m_monthly,
                   aspect=2.25, ylab="Temperature",
                   ss.w = 11, ss.d = 1, fc.w = 91, fc.d = 1)
dev.off()

write.csv(D30m.STL[[3]], paste(dataOUTDIR, "D30m_monthly.csv", sep="/"))
write.csv(D30m.STL[[1]], paste(dataOUTDIR, "D30m_components.csv", sep="/"))

pdf(paste(figDIR, "D60mSTL.pdf", sep="/"), height=4, width=5)
D60m.STL <- stl.rfs2(data=D60m_monthly,
                   aspect=2.25, ylab="Temperature",
                   ss.w = 11, ss.d = 1, fc.w = 91, fc.d = 1)
dev.off()

write.csv(D60m.STL[[3]], paste(dataOUTDIR, "D60m_monthly.csv", sep="/"))
write.csv(D60m.STL[[1]], paste(dataOUTDIR, "D60m_components.csv", sep="/"))

pdf(paste(figDIR, "D75mSTL.pdf", sep="/"), height=4, width=5)
D75m.STL <- stl.rfs2(data=D75m_monthly,
                   aspect=2.25, ylab="Temperature",
                   ss.w = 11, ss.d = 1, fc.w = 91, fc.d = 1)
dev.off()

write.csv(D75m.STL[[3]], paste(dataOUTDIR, "D75m_monthly.csv", sep="/"))
write.csv(D75m.STL[[1]], paste(dataOUTDIR, "D75m_components.csv", sep="/"))

pdf(paste(figDIR, "D100STL.pdf", sep="/"), height=4, width=5)
D100.STL <- stl.rfs2(data=D100_monthly,
                   aspect=2.25, ylab="Temperature",
                   ss.w = 11, ss.d = 1, fc.w = 91, fc.d = 1)
dev.off()

write.csv(D100.STL[[3]], paste(dataOUTDIR, "D100_monthly.csv", sep="/"))
write.csv(D100.STL[[1]], paste(dataOUTDIR, "D100_components.csv", sep="/"))

pdf(paste(figDIR, "D110STL.pdf", sep="/"), height=4, width=5)
D110.STL <- stl.rfs2(data=D110_monthly,
                   aspect=2.25, ylab="Temperature",
                   ss.w = 11, ss.d = 1, fc.w = 91, fc.d = 1)
dev.off()

write.csv(D110.STL[[3]], paste(dataOUTDIR, "D110_monthly.csv", sep="/"))
write.csv(D110.STL[[1]], paste(dataOUTDIR, "D110_components.csv", sep="/"))

pdf(paste(figDIR, "D120STL.pdf", sep="/"), height=4, width=5)
D120.STL <- stl.rfs2(data=D120_monthly,
                   aspect=2.25, ylab="Temperature",
                   ss.w = 11, ss.d = 1, fc.w = 91, fc.d = 1)
dev.off()

write.csv(D120.STL[[3]], paste(dataOUTDIR, "D120_monthly.csv", sep="/"))
write.csv(D120.STL[[1]], paste(dataOUTDIR, "D120_components.csv", sep="/"))

pdf(paste(figDIR, "D140STL.pdf", sep="/"), height=4, width=5)
D140.STL <- stl.rfs2(data=D140_monthly,
                   aspect=2.25, ylab="Temperature",
                   ss.w = 11, ss.d = 1, fc.w = 91, fc.d = 1)
dev.off()

write.csv(D140.STL[[3]], paste(dataOUTDIR, "D140_monthly.csv", sep="/"))
write.csv(D140.STL[[1]], paste(dataOUTDIR, "D140_components.csv", sep="/"))
```